Identifying and correcting multiple sources of misspecification\\in GWAS summary statistics for prediction\\---~OR~---\\Reducing misspecifications and improving tagging\\for better polygenic scores

### 2021-08-31 paper/paper-misspec.tex
start discu

@@

@@

We design simulations where variants have different GWAS sample sizes, which is often the case when meta-analyzing GWAS from multiple cohorts without the same genome coverage.
Using 40,000 variants from chromosome 22 (Methods), we simulate quantitative phenotypes with a heritability of 20\% and 2000 causal variants.
For half of the variants, we use 100\% of 300,000 individuals for GWAS, but use only 80\% for one quarter of variants and 60\% for the remaining quarter. 
We then run C+T, LDpred2-inf, LDpred2(-grid), LDpred2-auto, lassosum, lassosum2, PRS-CS and SBayesR by using either the true per-variant GWAS sample sizes, the total sample size, or imputed sample sizes. 
However, we do not show results for SBayesR because it always diverged and for PRS-CS because the overlap with the LD reference they provide is too small.
Averaged over 10 simulations, when providing true per-variant GWAS sample sizes, squared correlations between the polygenic scores and the simulated phenotypes are of 0.123 for C+T, 0.161 for lassosum, 0.169 for lassosum2, 0.159 for LDpred2(-grid), 0.140 for LDpred2-auto, and 0.141 for LDpred2-inf (Figure \ref{fig:simu-misN}).
Results when using imputed (instead of true) sample sizes are quite similar.
Note that C+T does not use this sample size information.
When using the total GWAS sample size instead of the per-variant sample sizes, predictive performance slightly decreases to 0.157 for lassosum and to 0.163 for lassosum2, but dramatically decreases for LDpred2 with new values of 0.134 for LDpred2-grid, 0.119 for LDpred2-auto, and 0.123 for LDpred2-inf (Figure \ref{fig:simu-misN}).
Therefore, this extreme simulation scenario shows that LDpred2 can be sensitive to GWAS sample size misspecification, whereas lassosum (and lassosum2) seems little affected by this.

@@

Using 40,000 variants from chromosome 22, we simulate quantitative phenotypes assuming an heritability of 20\% and 2000 causal variants using the ``true'' dataset (cf.\ Methods section ``Data for simulations'').
We compute GWAS summary statistics from the dosage dataset and use these summary statistics to run LDpred2 and lassosum2 with either no correction of the summary statistics, or with one of the three corrections described above.
The LD reference used by LDpred2 and lassosum2 is computed from the validation set using the dataset with the ``true'' genotypes.
For lassosum2 and LDpred2(-grid) which tune parameters using the validation set, correcting for imputation quality slightly improves predictive performance in these simulations (Figure \ref{fig:simu-info}).
However, correcting for imputation quality can dramatically improves predictive performance for LDpred2-auto. 
Moreover, new additions for robustness introduced before, namely LDpred2-low-h2, LDpred2-auto-rob, and forming independent blocks in the LD matrix also improve predictive performance for all corrections (Figure \ref{fig:simu-info}).

@@

\begin{figure}[h]
	\centerline{\includegraphics[width=0.95\textwidth]{brca_onco_qc}}
	\caption{Standard deviations inferred from the OncoArray breast cancer GWAS summary statistics versus the ones inferred from the reported GWAS allele frequencies. Only 100,000 HapMap3 variants are represented, at random.}
	\label{fig:brca_onco_qc}
\end{figure}

Therefore, when an accurate imputation INFO score is provided, this enables both to correct for it in the QC step we propose to possibly uncover problems with the summary statistics, and when building polygenic scores to improve prediction.
We try this for the two breast cancer summary statistics. We first compare the standard QC proposed in \cite{prive2020ldpred2} (which ends up filtering on MAF here, which we call ``qc1''). We then also filter out the two long-range LD regions of chromosome 6 and 8 for the OncoArray summary statistics and remove around 500 variants when filtering on differences of MAFs between summary statistics and the validation dataset (``qc2'').
As for the correction for the INFO scores, we use the first correction, ``sqrt\_info'', which is simple because it is equivalent to post-processing PGS effects by multiplying them by $\sqrt{\text{INFO}}$.
Although results for both QC used are very similar, correcting for the INFO score slightly improves prediction when deriving polygenic scores based on iCOGS summary statistics (Figure \ref{fig:res_brca}). 
All other improvements introduced before have little or no effect here [DEPRESSING; SHOULD REWORD?].

@@

\subsection{Larger sets of variants and other phenotypes}

We use external summary statistics for which INFO scores are reported (Table \ref{tab:sumstats}); they all have a very high mean INFO score (larger than 0.94), except for BRCA-iCOGS (0.841) and T1D-affy (0.885).
QC plots comparing standard deviations usually look good (after the INFO score correction), except for coronary artery disease (CAD) summary statistics (Figures \ref{fig:brca_icogs_qc}-\ref{fig:t1d_illu_qc}).
We show previous results as figure \ref{fig:res_brca} for other phenotypes in figures \ref{fig:res_cad}-\ref{fig:res_t1d}; most changes introduced before have little or no impact on the predictions, except for the second quality control performed for CAD when using LDpred2-auto (Figure \ref{fig:res_cad}).
We then show results with two new sets of variants (Methods) in figures \ref{fig:res_brca_all}-\ref{fig:res_t1d_all}.
[WHAT TO SAY ABOUT THIS? MAXTAG GOOD FOR SOME TRAITS, CLUMP NOT GREAT. PUT ONE FIGURE IN MAIN TEXT?]

@@

@@

@@

We can then impute $n_j$ from equation \eqref{eq:approx-sd-lin} using
\begin{equation}\label{eq:imputeN}
n_j \approx \dfrac{\text{var}(y) / \text{var}(G_j) - \hat{\gamma}_j^2}{\text{se}(\hat{\gamma}_j)^2} ~,
\end{equation}
and impute $n_j^\text{eff}$ from equation \eqref{eq:approx-sd-log} using
\begin{equation}\label{eq:imputeN}
n_j^\text{eff} \approx \dfrac{4 / \text{var}(G_j) - \hat{\gamma}_j^2}{\text{se}(\hat{\gamma}_j)^2} ~.
\end{equation}
In practice, we also bound this estimate to be between $0.5 \cdot N$ and $1.1 \cdot N$, where $N$ is the global sample size.

>>>>

[SHOULD ADD THE LDAK-ALPHA IN THIS WORK? MAYBE LATER -> IN A THIRD VERSION]

@@

[WHY CHOOSED TO FOCUS ON LDPRED AND LASSOSUM -> COPY PREVIOUS INTRO + MORE CONTROL ON THE IMPLEMENTATION]

[WHAT I WANTED TO INVESTIGATE]

@@

We design simulations where variants have different GWAS sample sizes, which is often the case when meta-analyzing GWAS from multiple cohorts without the same genome coverage.
Using 40,000 variants from chromosome 22 (Methods), we simulate quantitative phenotypes with a heritability of 20\% and 2000 causal variants.
For half of the variants, we use 100\% of 300,000 individuals for GWAS, but use only 80\% for one quarter of variants and 60\% for the remaining quarter. 
We then run C+T, LDpred2-inf, LDpred2(-grid), LDpred2-auto, lassosum, lassosum2 (Methods), PRS-CS and SBayesR \cite[]{prive2019making,prive2020ldpred2,mak2017polygenic,ge2019polygenic,lloyd2019improved} by using either the true per-variant GWAS sample sizes, the total sample size, or imputed sample sizes. 
However, we do not show results for SBayesR because it always diverged and for PRS-CS because the overlap with the LD reference they provide is too small.
Averaged over 10 simulations, when providing true per-variant GWAS sample sizes, squared correlations between the polygenic scores and the simulated phenotypes are of 0.123 for C+T, 0.161 for lassosum, 0.169 for lassosum2, 0.159 for LDpred2(-grid), 0.140 for LDpred2-auto, and 0.141 for LDpred2-inf (Figure \ref{fig:simu-misN}).
Results when using imputed (instead of true) sample sizes are quite similar.
Note that C+T does not use this sample size information.
When using the total GWAS sample size instead of the per-variant sample sizes, predictive performance slightly decreases to 0.157 for lassosum and to 0.163 for lassosum2, but dramatically decreases for LDpred2 with new values of 0.134 for LDpred2-grid, 0.119 for LDpred2-auto, and 0.123 for LDpred2-inf (Figure \ref{fig:simu-misN}).
Therefore, this extreme simulation scenario shows that LDpred2 can be sensitive to GWAS sample size misspecification, whereas lassosum (and lassosum2) seems little affected by this.

@@

Using 40,000 variants from chromosome 22, we simulate quantitative phenotypes assuming an heritability of 20\% and 2000 causal variants using the ``true'' dataset (cf.\ Methods section ``Data for simulations'').
We compute GWAS summary statistics from the dosage dataset and use these summary statistics to run LDpred2 and lassosum2 with either no correction of the summary statistics, or with one of the three corrections described above.
The LD reference used by LDpred2 and lassosum2 is computed from the validation set using the dataset with the ``true'' genotypes.
For lassosum2 and LDpred2(-grid) which tune parameters using the validation set, correcting for imputation quality slightly improves predictive performance in these simulations (Figure \ref{fig:simu-info}).
However, correcting for imputation quality can dramatically improve predictive performance for LDpred2-auto. 
Moreover, new additions for robustness introduced before, namely LDpred2-low-h2, LDpred2-auto-rob, and forming independent blocks in the LD matrix also improve predictive performance for all corrections (Figure \ref{fig:simu-info}).

@@

\begin{figure}[h]
	\centerline{\includegraphics[width=0.95\textwidth]{brca_onco_qc}}
	\caption{Standard deviations inferred from the OncoArray breast cancer GWAS summary statistics (Equation \ref{eq:approx-sd-log}) versus the ones inferred from the reported GWAS allele frequencies ($\sqrt{2 f (1 - f)}$). Only 100,000 HapMap3 variants are represented, at random.}
	\label{fig:brca_onco_qc}
\end{figure}

Therefore, when an accurate imputation INFO score is provided, this enables both to correct for it in the QC step we propose to possibly uncover problems with the summary statistics, and when building polygenic scores to improve prediction.
We apply this correction to the two breast cancer summary statistics. We first compare the standard QC proposed in \cite{prive2020ldpred2} (which ends up filtering on MAF here, which we call ``qc1''). We then also filter out the two long-range LD regions of chromosome 6 and 8 for the OncoArray summary statistics and remove around 500 variants when filtering on differences of MAFs between summary statistics and the validation dataset (``qc2'').
As for the correction for the INFO scores, we use the first correction, ``sqrt\_info'', which is simple because it is equivalent to post-processing PGS effects by multiplying them by $\sqrt{\text{INFO}}$.
Although results for both QC used are very similar, correcting for the INFO score slightly improves prediction when deriving polygenic scores based on iCOGS summary statistics (Figure \ref{fig:res_brca}). 
All other improvements introduced before have little or no effect here, probably because misspecifications are much smaller than in the simulations.

@@

\subsection{Other phenotypes and larger sets of variants}

We use external summary statistics for which INFO scores are reported (Table \ref{tab:sumstats}); they all have a very high mean INFO score (larger than 0.94), except for BRCA-iCOGS (0.841) and T1D-affy (0.885).
QC plots comparing standard deviations usually look good (after the INFO score correction), except for coronary artery disease (CAD) summary statistics (Figures \ref{fig:brca_icogs_qc}-\ref{fig:t1d_illu_qc}).
We show previous results as figure \ref{fig:res_brca} for other phenotypes in figures \ref{fig:res_cad}-\ref{fig:res_t1d}; most changes introduced before have little or no impact on the predictions, except for the second quality control performed for CAD when using LDpred2-auto (Figure \ref{fig:res_cad}).

We then introduce two new sets of variants as possible replacement of HapMap3 variants (Methods).
These sets include more variants (more than 2M), therefore possibly of lower quality on average. 
We show results with these two new sets of variants (Methods) in figures \ref{fig:res_brca_all}-\ref{fig:res_t1d_all}.
Using the ``maxtag'' set generally provides larger predictive performance than using HapMap3 variants, especially when predicting prostate cancer (Figure \ref{fig:res_prca_all}).

@@

Here we have investigated misspecifications in GWAS summary statistics, including when they are derived from imputed data, and when the information on sample size per variant is different for all variants because of meta-analysis.
Using LDpred2 models \cite[]{prive2020ldpred2} and our reimplementation of the lassosum model \cite[]{mak2017polygenic} for deriving polygenic scores, we have introduced some changes to possibly make these models more robust to these misspecifications.
Previously, we proposed a quality control based on comparing standard deviations derived from GWAS summary statistics with the ones from a reference panel \cite[]{prive2020ldpred2}.
Here we show that we can refine this quality control and almost get perfect matches by deriving standard deviations from reported allele frequencies in the summary statistics (instead of from the reference panel, which can now be used to compare allele frequencies), and by also correcting for the INFO score (e.g.\ see figure \ref{fig:brca_onco_qc}).
Using this refined QC, we were able to identify a potential issue with how principal components were derived in a set of breast cancer summary statistics. Fortunately this has practically no effect on the predictive performance of the derived polygenic scores.

Note that, in this study, we use summary statistics for which this information (INFO scores and allele frequencies) is available, but this is far from being the case for all GWAS summary statistics. 
We acknowledge that, in the case of a meta-analysis from multiple studies, providing a single INFO score may not be possible; would a weighted averaged INFO score work?
Nevertheless, this quality control could be easily performed within each study before pooling results, to make sure that summary statistics have the best possible quality for follow-up analyses such as deriving polygenic scores.
Finally, another information, the effect sample size per variant, is often missing from GWAS summary statistics. 
We recall that if some studies of a meta-analysis have an imbalanced number of cases and controls, the global effective sample size should not be computed from the total numbers of cases and controls overall, but instead from the sum of each effective sample size.
Indeed, take the extreme example of meta-analyzing two studies, one with 1000 cases and 0 controls, and another one with 0 cases and 1000 controls, then the effective sample size of the meta-analysis is 0, not 2000.
Again, we note that an overestimated sample size can be detected from the QC plot we propose (when the slope is less than 1) for case-control studies using logistic regression; otherwise the standard deviation of the phenotype is also needed (Equation \ref{eq:approx-sd-lin}).

[TODO: FINISH]

@@

[BOTH DISAPOINTING BUT REASSURING]

@@

We can then impute $n_j$ from equation \eqref{eq:approx-sd-lin} using
\begin{equation}\label{eq:imputeN}
n_j \approx \dfrac{\text{var}(y) / \text{var}(G_j) - \hat{\gamma}_j^2}{\text{se}(\hat{\gamma}_j)^2} ~,
\end{equation}
and impute $n_j^\text{eff}$ from equation \eqref{eq:approx-sd-log} using
\begin{equation}\label{eq:imputeNeff}
n_j^\text{eff} \approx \dfrac{4 / \text{var}(G_j) - \hat{\gamma}_j^2}{\text{se}(\hat{\gamma}_j)^2} ~.
\end{equation}
In practice, we also bound this estimate to be between $0.5 \cdot N$ and $1.1 \cdot N$, where $N$ is the global sample size.
